Landform boundary effects on Holocene forager landscape use in arid South Australia

### 2017-02-04 analysis/paper/paper.Rmd
make citations and references more consistent, remove data files from repo

A key step in understanding past human use of landscapes is investigating how people adapted their behaviors to the environmental variations they encountered. When analysing the residues of these behaviours, archaeologists often depend on comparing modal attributes of archaeological sites between discrete patches of landscapes (e.g. Barton 2003). Variations in human behavior at landscape boundaries are more challenging to investigate, despite their importance in understanding sensitivity to landscape differences. Previous work on past human behaviors at ecological interfaces has focused on regions of highly contrasting biomes. For example, Epp (1985) found that archaeological sites were more concentrated in an ecotone (i.e. ecological transition area) where plains grasslands meet the boreal forest in Saskatchewan, compared to within those biomes. Epp’s work is typical of much early archaeological work on ecotones, with claims that ecotonal areas contain higher populations and diversities of flora and fauna (Struever 1968, Glassow and Leone 1972). This increased variety and density of plants and animals, often described as an ‘edge effect’, has been a focus of archaeological research. For example, it was claimed by Harris et al. (1969) to be a causal factor in the emergence of agriculture. This edge effect was also extended to human culture, with Gummerman and Johnson (1971) arguing for increased human cultural diversity in ecotonal zones of central Arizona. In Australia edge effects in landscape resources and their exploitation have also been discussed. For instance, Hughes and Hiscock (1982) noted apparent edge effects in the concentration of stone artefact scatters along landscape boundaries in arid northern South Australia.

These uses of the ecotone and edge effect concepts declined in the archaeological literature after several critiques pointed out that archaeologists had been overly simplistic in their use of many ecological concepts (Rhoades 1978, Hardesty 1980, King and Graham 1981). Problems with these previous approaches include ignoring controversies among biologists about the definition of ecotones, unjustified assumptions about the uniformity of all ecotones, and unverified assumptions about edge effects always increasing the amount of food species available to humans. Alternative approaches to investigating landform boundary effects can be found in the work of Kimes et al. (1982), who used distribution of artefacts to see how culture group areas related to landform boundaries. However, much recent work on archaeological settlement patterns of hunter-gatherers has eschewed boundaries in favour of analysing straight line distances from key resources such as raw materials for stone artefacts, as these can be easier to measure and interpret than landscape boundaries (Daniel 2001; Jones, Beck et al. 2003). Here we pursue a new approach to understanding boundaries by focusing on subtle variations in landscape characteristics within a biome, rather than between them. By focusing on continuous variation, rather than step-wise shifts in the archaeological record across ecological interfaces, we introduce a new approach to studying behaviors of prehistoric forager groups at interface zones.

In this study we use the results from a sample of hand-held GIS-enabled archaeological site recordings to ask if it is possible to observe gradients of change in archaeological materials over distance to understand forager choices of the optimum locations for their activities in respect to environmental boundaries. We predict that prehistoric foragers were sensitive to subtle differences in resource structure as gradients across the landscape. This approach may be contrasted with concepts of linear boundaries with step-wise shifts in habitability. The step-wise approaches are useful because they make comparative landscape analysis more tractable in many contexts and at different scales. For example, the spatial scales of much previous work on prehistoric arid zone foragers has ranged from continental (Williams, Ulm et al. 2013; Bird et al 2016) to ecological regions (Holdaway, Douglass et al. 2013). In this paper, we introduce a novel focus on human behavior in the arid zone: the interface between two types of subtly different landscapes. We show that prehistoric forager groups' decisions about how they occupied the study area were sensitive to subtle differences in local landscape characteristics.

Our work is motivated by previous archaeological work on cultural landscapes of foragers in Australian arid and semi-arid zones that emphasizes the relationship between settlement organization and both exploitable resources and prominent landscape features such as dunes, occurrences of flakeable rocks, pans, springs and salt lakes (Hughes and Hiscock 1982, Veth 1993, Napton and Greathouse 1996, Holdaway 1998, Barton 2003, Smith 2006, Fanning, Holdaway et al. 2008, Smith, Williams et al. 2008, Veth, Smith et al. 2008, Williams, Ulm et al. 2013, Davies, Holdaway et al. 2015; Gould and Saggers 1985). The archaeological signature of landuse on these landscapes has typically been the distribution and composition of surface stone artefact scatters, which are by far the most abundant type of site. Sites in these landscapes are typically small, dense and discrete concentrations of stone artefacts that represent temporary, short-term activity areas.

@@

The Roxby dunefield is in the driest part of the Australian arid zone (Figure \@ref(fig:overviewmap)). Archaeological material is found across the broader region and partly reflects the long period of human occupation, the wide range of locally available resources and the extremely low rates of sediment movement that might otherwise obscure surface artefacts. Optically stimulated luminescence (OSL) dating of sediments underling buried artefacts in several sites demonstrate episodic human use of the area from the late Pleistocene deglacial period (from around 19 ka, Hughes et al. 2014b; Sullivan, Field et al. 2012; Sullivan et al. 2014!!b). No organic materials have been found at any of the open sites or in the excavated deposits. Distinctive implement forms such as backed artefacts, unifacial points and tula adzes at many sites in the study area indicate a mid to late Holocene age for the main period of land use and for most of the surface artefact clusters recorded during the survey (Hughes and Hiscock 2005). A study of 78 OSL ages from geological contexts in the study area has shown that the mid to late Holocene period was a time of broad-scale dune stability (Hughes et al. 2014b). This means that site preservation and visibility is not strongly controlled by dune movement. In 2013 we returned to sites that had been recorded 25 years earlier and found them to be in the same state as originally recorded, indicating that site visibility has changed little over 25 years.

@@

The Roxby dunefield has been the target of mineral exploration and mining since the 1980s. The extensive records of archaeological sites near the Olympic Dam mine site in the duenfield have been created during research-orientated consultancy work related to this exploration and mining activity. In 1980 Hughes, Hiscock and colleagues commenced a range of archaeological investigations for the then-proposed mining project. They used terrain patterns based on combining categories of different landforms and surface geology (Figure \@ref(fig:mapofstudyarea1)) to develop an environmentally-based model to predict the nature and distribution of archaeological sites (Hughes and Hiscock 1982). The landscape consists of ancient stony plateau (or tableland) surfaces partly overlain by fields of longitudinal east-west dunes. Many of the interdune corridors (or swales) and extensive flat to gently sloping plateau surfaces without dunes are covered by a lag of gravels and cobbles referred to as gibber. Water sources include ephemeral small lakes, claypans and canegrass-vegetated swamps which are subject to high evaporation rates, with no free water available for most of the year. In this region vegetation patterns were also determined by the presence or absence of sand, gibber surfaces and water bodies (Badman 1999). This means that there are long-term relationships between the landscape units employed in the analysis of Hughes and Hiscock (1982), and plant and animal resources exploited by forager communities in the past.

@@

Hughes and Hiscock’s archaeological model has been described in detail in previous publications (Hughes et al. 2011; Hughes and Hiscock 1982; Hughes and Sullivan 1984). In brief, the model details how the location and character of archaeological sites are influenced by proximity to fresh water, sand on which to camp, and locally available raw materials for flaking stone artefacts. The model results in a series of predictions about the archaeological record expected for each of six landform types (Table \@ref(tab:table1)) and five geological regimes (Table \@ref(tab:table2)). These predictions specify the frequency of sites and the density and diversity of artefacts were produced for each major landscape unit.

These descriptions were initially tested in the 1980s and 1990s by archaeological surveys of the mining area and infrastructure corridors. The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites (Hiscock and Hughes 1983, Hiscock 1989). In 2007 a proposed expansion of the existing mine triggered the need for an intensive archaeological survey of an additional area of 515 km^2^ surrounding the existing mine, to be completed by the end of 2009. The data collected during this expansion phase is the basis for this study (Figure \@ref(fig:mapofstudyarea2)).

@@

Between 2007 and 2009 teams of archaeologists and trainees from local Aboriginal groups walking across the entire area recorded more than 17,000 archaeological sites (Sullivan et al. 2014a). In this study we draw on a sample of `r total_number_of_sites` sites from a 353 km^2^ area within the Roxby dunefield (Figure \@ref(fig:mapofstudyarea2)). The sites consist of stone artefact clusters (`r artefact_scatters`%), artefact clusters with knapping floors (`r as_kf`%), knapping floors (`r knapping_floors`%) and quarries (`r quarries`%) (Hughes, Hiscock et al. 2011). Extended descriptions of site definitions and survey methods have previously been presented in Hughes et al. (2011) and Sullivan et al. 2014a.

To collect field data, we used a mobile geographic information system (GIS) on handheld computers to record both background environmental and cadastral information and archaeological information. Our equipment was similar to what is commonly used during field survey (McPherron and Dibble 2003, Bevan and Conolly 2004, Tripcevich 2004, 2004, Given and Hyla 2006, Wagtendonk and De Jeu 2007, de la Vega and Agulla 2010, Tripcevich and Wernke 2010, Fei 2011, Scianna and Villa 2011, Traviglia 2011, Fee, Pettegrew et al. 2013, Newhard, Levine et al. 2013, Cascalheira, Gonçalves et al. 2014, Banning and Hitchings 2015). Our mobile GIS configuration files, recording forms, and scripts are archived online at <http://doi.org/10.5281/zenodo.11833>. A detailed description of our specific hardware, software, and justifications for our choices is presented in our SOM.

@@

To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2016), we include the entire R code used for all the analysis and visualizations contained in this paper in our SOM at <http://dx.doi.org/10.17605/OSF.IO/RMKGE>. Also in this version-controlled compendium are the raw data for all the tests and visualizations reported here, as well as additional regression diagnostics. All of the figures, tables and statistical test results presented here can be independently reproduced with the code and data in this repository. In our SOM our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2016).

@@

The data from the Roxby dunefield allow a test of the predictions of the landuse model developed for the region in the early 1980s (Hughes, Hiscock et al. 2011). The first prediction tested here is that there was differential use of terrain pattern areas within dunefields. More specifically, we predicted that areas of widely-spaced longitudinal sand dunes were intensely occupied, compared with areas containing more closely-spaced dunes. This difference is anticipated because of the greater range and density of animal foods, stone for making artefacts, and water sources in places where dunes are more widely separated.

@@

# dot-plot
library(viridis)
ggplot(bound.sc.m, 
       aes(x=distance, 
           y=`density of sites`,
           shape = `Terrain Pattern`,
           colour = `Terrain Pattern`)) + 
  geom_point(size = 5)  +
  geom_line(aes(group = `Terrain Pattern`)) +
  theme_bw(base_size = text_size) +
  coord_flip() +
  scale_color_viridis(discrete = TRUE) +
  theme(axis.title.y = element_text(angle = 90)) + 
  xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
  ylab("z-score of site density in buffer area")  +
  labs(shape="Terrain\nPattern", colour="Terrain\nPattern")

@@

While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS. This is because the rate of recording is strongly predicted by the density of sites (F(1, `r mobile_gis_efficiency_lm_df`) = `r mobile_gis_efficiency_lm_F`, R^2^~adj~ = `r mobile_gis_efficiency_lm_r_squared`, p = `r mobile_gis_efficiency_lm_p_value`). To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data, which resulted in a rate of site recording that can by predicted by `r mobile_gis_efficiency_lm_2007_2009_inter` (`r mobile_gis_efficiency_lm_2007_2009_inter_error`) + `r mobile_gis_efficiency_lm_2007_2009_est`(site density) (`r mobile_gis_efficiency_lm_2007_2009_est_error`) +  (standard errors in parentheses). This model gave a good fit (R^2^~adj~ = `r mobile_gis_efficiency_lm_2007_2009_r_sq_adj`, p = `r mobile_gis_efficiency_lm_2007_2009_p_value`), and was then used to predict the site recording rate using a mobile GIS given the density of 19 sites/km^2^ observed in the pre-2007 data. The resulting rate is `r resulting_rate` sites/person/day, which is what the model predicts if a mobile GIS had been used to record sites in an area with 19 sites/km^2^. The difference of `r difference_in_rates_pre_2007_and_2007_2009` sites/person/day between the actual pre-2007 average rate of `r rate_for_pre_2007` and the predicted average rate of `r resulting_rate` is most likely due to the use of mobile GIS. This represents a `r percentage_increase`% increase in the site recording rate, although we caution that this may be an overestimate due to overfitting from the relatively small number of data points in our efficiency analysis. Nevertheless, this demonstrates an advantage for using mobile GIS in a context where data collection was intensive, data volume was high, a highly structured data collection form was used, and the start-up, purchasing, and training costs could be amortized over multiple years of data collection.

@@

The substantive archaeological contribution of this study has been to show that when choosing habitations, prehistoric people in the Roxby dunefield were sensitive to relatively small variations in the landscape, such as the spacing of dunes. The implication of this finding is that even in relatively homogeneous landscapes, such as the desert investigated here, there are gradient-like patterns in the distribution of archaeological sites the reveal how people adapted to the landscape (cf. Attwell and Fletcher 1987). These gradients allow for more sophisticated interpretations of prehistoric behaviour beyond the obvious attraction of prominent landscape features (Wilson 2007, Kowalewski 2008: 233). Given a large enough dataset, such as the one presented here, these patterns can be quickly quantified using a geographical information system and explained with relatively simple behavioral ecological models. We hope that this data-intensive approach to using gradients of changes in archaeological site attributes across the landscape might suggest a new direction for investigating past human behaviour at ecological interfaces, avoiding the problematic assumptions of previous work on edge effects.

@@

@@

de la Vega, S. G.-D. and S. O. Agulla (2010). Using mobile GIS. for Field Digital Data Acquisition (FDA) in archaeological field survey. Geoinformation Technologies for Geo-Cultural Landscapes: European Perspectives. A. Vassilopoulos, N. Evelpidou, O. Bender and A. Krek: 35-46.
Dunnell, R. C. and W. S. Dancey (1983). "The Siteless Survey: A Regional Scale Data Collection Strategy." Advances in Archaeological Method and Theory 6: 267-287.

@@

@@

Napton, L. K. and E. A. Greathouse (1996). Archaeological investigations in the Simpson Desert, Northern Territory. Archaeology of Northern Australia. P. Veth and P. Hiscock. Brisbane, University of Queensland. TEMPUS Volume 4: 106-136.

@@

@@

Sullivan, M., P. Hiscock, and P. Hughes (2014) Three Scales: GIS, GPS and Digital Site and Data Recording Technology in Archaeological Salvage at Olympic Dam in Arid South Australia. Journal of the Anthropological Society of South Australia 38: 85-107.

Sullivan, M., P. Hughes, A. Mosig Way and N. Spooner (2014).	Prehistoric mining at Olympic Dam in arid South Australia. Archaeology in Oceania 49(1):43-55.

@@

Thomas, D. H. (1975). Nonsite sampling in archaeology: up the creek without a site. Sampling in archaeology. J. W. Mueller. Tucson, AZ, University of Arizona Press: 61-81.
Traviglia, A. (2011). Integrated Archaeological Investigations for the Study of the Greater Aquileia Area. The New Technologies for Aquileia, Aquileia, Italy.

Tripcevich, N. (2004). "Flexibility by Design: How Mobile GIS Meets the Needs of Archaeological Survey." Cartography and Geographic Information Science 31(3): 137-151.

Tripcevich, N. (2004). "Mobile GIS in archaeological survey." The SAA Archaeological Record(May).

>>>>

A key step in understanding past human use of landscapes is investigating how people adapted their behaviors to the environmental variations they encountered. When analysing the residues of these behaviours, archaeologists often depend on comparing modal attributes of archaeological sites between discrete patches of landscapes (e.g. Barton 2003). Variations in human behavior at landscape boundaries are more challenging to investigate, despite their importance in understanding sensitivity to landscape differences. Previous work on past human behaviors at ecological interfaces has focused on regions of highly contrasting biomes. For example, Epp (1985) found that archaeological sites were more concentrated in an ecotone (i.e. ecological transition area) where plains grasslands meet the boreal forest in Saskatchewan, compared to within those biomes. Epp’s work is typical of much early archaeological work on ecotones, with claims that ecotonal areas contain higher populations and diversities of flora and fauna (Struever 1968; Glassow and Leone 1972). This increased variety and density of plants and animals, often described as an ‘edge effect’, has been a focus of archaeological research. For example, it was claimed by Harris et al. (1969) to be a causal factor in the emergence of agriculture. This edge effect was also extended to human culture, with Gummerman and Johnson (1971) arguing for increased human cultural diversity in ecotonal zones of central Arizona. In Australia edge effects in landscape resources and their exploitation have also been discussed. For instance, Hughes and Hiscock (1982) noted apparent edge effects in the concentration of stone artefact scatters along landscape boundaries in arid northern South Australia.

These uses of the ecotone and edge effect concepts declined in the archaeological literature after several critiques pointed out that archaeologists had been overly simplistic in their use of many ecological concepts (Rhoades 1978; Hardesty 1980; King and Graham 1981). Problems with these previous approaches include ignoring controversies among biologists about the definition of ecotones, unjustified assumptions about the uniformity of all ecotones, and unverified assumptions about edge effects always increasing the amount of food species available to humans. Alternative approaches to investigating landform boundary effects can be found in the work of Kimes et al. (1982), who used distribution of artefacts to see how culture group areas related to landform boundaries. However, much recent work on archaeological settlement patterns of hunter-gatherers has eschewed boundaries in favour of analysing straight line distances from key resources such as raw materials for stone artefacts, as these can be easier to measure and interpret than landscape boundaries (Daniel 2001; Jones et al. 2003). Here we pursue a new approach to understanding boundaries by focusing on subtle variations in landscape characteristics within a biome, rather than between them. By focusing on continuous variation, rather than step-wise shifts in the archaeological record across ecological interfaces, we introduce a new approach to studying behaviors of prehistoric forager groups at interface zones.

In this study we use the results from a sample of hand-held GIS-enabled archaeological site recordings to ask if it is possible to observe gradients of change in archaeological materials over distance to understand forager choices of the optimum locations for their activities in respect to environmental boundaries. We predict that prehistoric foragers were sensitive to subtle differences in resource structure as gradients across the landscape. This approach may be contrasted with concepts of linear boundaries with step-wise shifts in habitability. The step-wise approaches are useful because they make comparative landscape analysis more tractable in many contexts and at different scales. For example, the spatial scales of much previous work on prehistoric arid zone foragers has ranged from continental (Williams et al. 2013; Bird et al 2016) to ecological regions (Holdaway et al. 2013). In this paper, we introduce a novel focus on human behavior in the arid zone: the interface between two types of subtly different landscapes. We show that prehistoric forager groups' decisions about how they occupied the study area were sensitive to subtle differences in local landscape characteristics.

Our work is motivated by previous archaeological work on cultural landscapes of foragers in Australian arid and semi-arid zones that emphasizes the relationship between settlement organization and both exploitable resources and prominent landscape features such as dunes, occurrences of flakeable rocks, pans, springs and salt lakes (Hughes and Hiscock 1982; Veth 1993; Napton and Greathouse 1996; Holdaway 1998; Barton 2003; Smith 2006; Fanning et al. 2008; Smith et al. 2008; Veth et al. 2008; Williams et al. 2013; Davies et al. 2015; Gould and Saggers 1985). The archaeological signature of landuse on these landscapes has typically been the distribution and composition of surface stone artefact scatters, which are by far the most abundant type of site. Sites in these landscapes are typically small, dense and discrete concentrations of stone artefacts that represent temporary, short-term activity areas.

@@

The Roxby dunefield is in the driest part of the Australian arid zone (Figure \@ref(fig:overviewmap)). Archaeological material is found across the broader region and partly reflects the long period of human occupation, the wide range of locally available resources and the extremely low rates of sediment movement that might otherwise obscure surface artefacts. Optically stimulated luminescence (OSL) dating of sediments underling buried artefacts in several sites demonstrate episodic human use of the area from the late Pleistocene deglacial period (from around 19 ka, Hughes et al. 2014b; Sullivan et al. 2012; Sullivan et al. 2014b). No organic materials have been found at any of the open sites or in the excavated deposits. Distinctive implement forms such as backed artefacts, unifacial points and tula adzes at many sites in the study area indicate a mid to late Holocene age for the main period of land use and for most of the surface artefact clusters recorded during the survey (Hughes and Hiscock 2005). A study of 78 OSL ages from geological contexts in the study area has shown that the mid to late Holocene period was a time of broad-scale dune stability (Hughes et al. 2014b). This means that site preservation and visibility is not strongly controlled by dune movement. In 2013 we returned to sites that had been recorded 25 years earlier and found them to be in the same state as originally recorded, indicating that site visibility has changed little over 25 years.

@@

The Roxby dunefield has been the target of mineral exploration and mining since the 1980s. The extensive records of archaeological sites near the Olympic Dam mine site in the dunefield have been created during research-orientated consultancy work related to this exploration and mining activity. In 1980 Hughes, Hiscock and colleagues commenced a range of archaeological investigations for the then-proposed mining project. They used terrain patterns based on combining categories of different landforms and surface geology (Figure \@ref(fig:mapofstudyarea1)) to develop an environmentally-based model to predict the nature and distribution of archaeological sites (Hughes and Hiscock 1982). The landscape consists of ancient stony plateau (or tableland) surfaces partly overlain by fields of longitudinal east-west dunes. Many of the interdune corridors (or swales) and extensive flat to gently sloping plateau surfaces without dunes are covered by a lag of gravels and cobbles referred to as gibber. Water sources include ephemeral small lakes, claypans and canegrass-vegetated swamps which are subject to high evaporation rates, with no free water available for most of the year. In this region vegetation patterns were also determined by the presence or absence of sand, gibber surfaces and water bodies (Badman 1999). This means that there are long-term relationships between the landscape units employed in the analysis of Hughes and Hiscock (1982), and plant and animal resources exploited by forager communities in the past.

@@

Hughes and Hiscock’s archaeological model has been described in detail in previous publications (Hughes et al. 2011; Hughes and Hiscock 1982). In brief, the model details how the location and character of archaeological sites are influenced by proximity to fresh water, sand on which to camp, and locally available raw materials for flaking stone artefacts. The model results in a series of predictions about the archaeological record expected for each of six landform types (Table \@ref(tab:table1)) and five geological regimes (Table \@ref(tab:table2)). These predictions specify the frequency of sites and the density and diversity of artefacts were produced for each major landscape unit.

These descriptions were initially tested in the 1980s and 1990s by archaeological surveys of the mining area and infrastructure corridors. The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites (Hiscock and Hughes 1983; Hiscock 1989). In 2007 a proposed expansion of the existing mine triggered the need for an intensive archaeological survey of an additional area of 515 km^2^ surrounding the existing mine, to be completed by the end of 2009. The data collected during this expansion phase is the basis for this study (Figure \@ref(fig:mapofstudyarea2)).

@@

Between 2007 and 2009 teams of archaeologists and trainees from local Aboriginal groups walking across the entire area recorded more than 17,000 archaeological sites (Sullivan et al. 2014a). In this study we draw on a sample of `r total_number_of_sites` sites from a 353 km^2^ area within the Roxby dunefield (Figure \@ref(fig:mapofstudyarea2)). The sites consist of stone artefact clusters (`r artefact_scatters`%), artefact clusters with knapping floors (`r as_kf`%), knapping floors (`r knapping_floors`%) and quarries (`r quarries`%) (Hughes et al. 2011). Extended descriptions of site definitions and survey methods have previously been presented in Hughes et al. (2011) and Sullivan et al. (2014a).

To collect field data, we used a mobile geographic information system (GIS) on handheld computers to record both background environmental and cadastral information and archaeological information. Our equipment was similar to what is commonly used during field survey (McPherron and Dibble 2003; Bevan and Conolly 2004; Tripcevich 2004a; 2004b; Given and Hyla 2006; Wagtendonk and De Jeu 2007; de la Vega and Agulla 2010; Tripcevich and Wernke 2010; Fei 2011; Scianna and Villa 2011, Traviglia 2011, Fee et al. 2013; Newhard et al. 2013; Cascalheira et al. 2014; Banning and Hitchings 2015). Our mobile GIS configuration files, recording forms, and scripts are archived online at <http://doi.org/10.5281/zenodo.11833>. A detailed description of our specific hardware, software, and justifications for our choices is presented in our SOM.

@@

To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2016), we include the entire R code used for all the analysis and visualizations contained in this paper in our SOM at <http://dx.doi.org/10.17605/OSF.IO/RMKGE>. Also in this version-controlled compendium are additional regression diagnostics. To respect the wishes of our stakeholders and protect the site locations from damage we have not made our data openly available. In our SOM our code is released under the MIT license, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2016).

@@

The data from the Roxby dunefield allow a test of the predictions of the landuse model developed for the region in the early 1980s (Hughes et al. 2011). The first prediction tested here is that there was differential use of terrain pattern areas within dunefields. More specifically, we predicted that areas of widely-spaced longitudinal sand dunes were intensely occupied, compared with areas containing more closely-spaced dunes. This difference is anticipated because of the greater range and density of animal foods, stone for making artefacts, and water sources in places where dunes are more widely separated.

@@

# dot-plot
library(viridis)
ggplot(bound.sc.m, 
       aes(x=distance, 
           y=`density of sites`,
           shape = `Terrain Pattern`,
           colour = `Terrain Pattern`)) + 
  geom_point(size = 5)  +
  geom_line(aes(group = `Terrain Pattern`)) +
  theme_bw(base_size = text_size) +
  coord_flip() +
  scale_color_viridis(discrete = TRUE) +
  theme(axis.title.y = element_text(angle = 90)) + 
  xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
  ylab("z-score of site density \nin buffer area")  +
  labs(shape="Terrain\nPattern", colour="Terrain\nPattern") +
  theme(legend.position = c(0.4, 1.01), 
        legend.justification = c(0.01, 1.05))

@@

While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS. This is because the rate of recording is strongly predicted by the density of sites (F(1, `r mobile_gis_efficiency_lm_df`) = `r mobile_gis_efficiency_lm_F`, R^2^~adj~ = `r mobile_gis_efficiency_lm_r_squared`, p = `r mobile_gis_efficiency_lm_p_value`). To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data, which resulted in a rate of site recording that can by predicted by `r mobile_gis_efficiency_lm_2007_2009_inter` (`r mobile_gis_efficiency_lm_2007_2009_inter_error`) + `r mobile_gis_efficiency_lm_2007_2009_est`(site density) (`r mobile_gis_efficiency_lm_2007_2009_est_error`) +  (standard errors in parentheses). This model gave a good fit (R^2^~adj~ = `r mobile_gis_efficiency_lm_2007_2009_r_sq_adj`, p = `r mobile_gis_efficiency_lm_2007_2009_p_value`), and was then used to predict the site recording rate using a mobile GIS given the density of 19 sites/km^2^ observed in the pre-2007 data. The resulting rate is `r resulting_rate` sites/person/day, which is what the model predicts if a mobile GIS had been used to record sites in an area with 19 sites/km^2^. The difference of `r difference_in_rates_pre_2007_and_2007_2009` sites/person/day between the actual pre-2007 average rate of `r rate_for_pre_2007` and the predicted average rate of `r resulting_rate` is most likely due to the use of mobile GIS. This represents a `r percentage_increase`% increase in the site recording rate, although we caution that this may be an overestimate due to overfitting from the relatively small number of data points in our efficiency analysis.

Nevertheless, this demonstrates an advantage for using mobile GIS in a context where data collection was intensive, data volume was high, a highly structured data collection form was used, and the start-up, purchasing, and training costs could be amortized over multiple years of data collection. Our gain of `r percentage_increase`% is within the range documented by other studies using mobile data collection technologies. For example, Poehler and Ellis (2012) report a 371% increase in efficiency using iPads at Pompeii, and Austin (2014) reports a 21-32% increase when using Open Data Kit (Hartung et al. 2010) to record skeletal remains. This wide variation in efficiency gains suggests that the magnitude of the gain is highly dependent on the context in which the mobile technologies are employed.

@@

The substantive archaeological contribution of this study has been to show that when choosing habitations, prehistoric people in the Roxby dunefield were sensitive to relatively small variations in the landscape, such as the spacing of dunes. The implication of this finding is that even in relatively homogeneous landscapes, such as the desert investigated here, there are gradient-like patterns in the distribution of archaeological sites the reveal how people adapted to the landscape (cf. Attwell and Fletcher 1987). These gradients allow for more sophisticated interpretations of prehistoric behaviour beyond the obvious attraction of prominent landscape features (Wilson 2007; Kowalewski 2008: 233). Given a large enough dataset, such as the one presented here, these patterns can be quickly quantified using a geographical information system and explained with relatively simple behavioral ecological models. We hope that this data-intensive approach to using gradients of changes in archaeological site attributes across the landscape might suggest a new direction for investigating past human behaviour at ecological interfaces, avoiding the problematic assumptions of previous work on edge effects.

@@

Austin, A. 2014 Mobilizing Archaeologists: Increasing the Quantity and Quality of Data Collected in the Field with Mobile Technology. Advances in Archaeological Practice 2(1):13-23.

@@

de la Vega, S. G.-D. and S. O. Agulla (2010). Using mobile GIS. for Field Digital Data Acquisition (FDA) in archaeological field survey. Geoinformation Technologies for Geo-Cultural Landscapes: European Perspectives. A. Vassilopoulos, N. Evelpidou, O. Bender and A. Krek: 35-46.

Dunnell, R. C. and W. S. Dancey (1983). "The Siteless Survey: A Regional Scale Data Collection Strategy." Advances in Archaeological Method and Theory 6: 267-287.

@@

Hartung, C., A. Lerer, Y. Anokwa, C. Tseng, W. Brunette and G. Borriello 2010 Open data kit: Tools to build information services for developing regions. Proceedings of the 4th ACM/IEEE International Conference on Information and Communication Technologies and Development: ACM.

@@

Marwick, B. (2016). Computational reproducibility in archaeological research: Basic principles and a case study of their implementation. Journal of Archaeological Method and Theory, 1-27, doi: 10.1007/s10816-015-9272-9
Napton, L. K. and E. A. Greathouse (1996). Archaeological investigations in the Simpson Desert, Northern Territory. Archaeology of Northern Australia. P. Veth and P. Hiscock. Brisbane, University of Queensland. TEMPUS Volume 4: 106-136.

@@

Poehler, E.E. and S.J.R. Ellis 2012 The 2011 Season of the Pompeii Quadriporticus Project: The Southern and Northern Sides. Fasti Online Documents and Research 249:1-12.

@@

@@

Sullivan, M., P. Hiscock, and P. Hughes (2014a) Three Scales: GIS, GPS and Digital Site and Data Recording Technology in Archaeological Salvage at Olympic Dam in Arid South Australia. Journal of the Anthropological Society of South Australia 38: 85-107.

Sullivan, M., P. Hughes, A. Mosig Way and N. Spooner (2014b). Prehistoric mining at Olympic Dam in arid South Australia. Archaeology in Oceania 49(1):43-55.

Thomas, D. H. (1975). Nonsite sampling in archaeology: up the creek without a site. Sampling in archaeology. J. W. Mueller. Tucson, AZ, University of Arizona Press: 61-81.

Traviglia, A. (2011). Integrated Archaeological Investigations for the Study of the Greater Aquileia Area. The New Technologies for Aquileia, Aquileia, Italy.

Tripcevich, N. (2004a). "Flexibility by Design: How Mobile GIS Meets the Needs of Archaeological Survey." Cartography and Geographic Information Science 31(3): 137-151.

Tripcevich, N. (2004b). "Mobile GIS in archaeological survey." The SAA Archaeological Record(May).
